\pkg{regsem}: Regularized Structural Equation Modeling

### 2017-03-24 draft1.tex
final first draft

\Abstract{
The \pkg{regsem} package in \proglang{R}, an implementation of
regularized structural equation modeling (RegSEM; Jacobucci, Grimm, and
McArdle 2016), was recently developed with the goal of incorporating
various forms of penalized likelihood estimation with a broad array of
structural equations models. The forms of regularization include both
the \textit{ridge} (Hoerl and Kennard 1970) and the least absolute
shrinkage and selection operator (\textit{lasso}; Tibshirani 1996),
along with sparser extensions. RegSEM is particularly useful for
structural equation models that have a small parameter to sample size
ratio, as the addition of penalties can reduce the complexity, thus
reducing the bias of the parameter estimates. The paper covers the
algorithmic details and an overview of the use of \pkg{regsem} with the
application of both factor analysis and latent growth curve models.
}

@@

As a simple running example, Figure 1 depicts a linear latent growth
curve model (e.g. Meredith and Tisak 1990) with four time points and ten
predictors for a simulated dataset. In this, a researcher may want to
test this model, but may only have a relatively small sample size
(e.g.~80). There are 29 estimated parameters in this model, resulting in
a estimated parameter to sample size ratio as far below even the most
liberal recommendations (e.g.~10:1 parameters to sample size; Kline
2015). In lieu of finding additional respondents, reducing the number of
parameters estimated is one effective strategy for reducing bias.
Specifically, the 20 estimated regressions from \textit{c1-c10} could be
reduced to a number that makes the ratio of the parameters estimated to
sample size more reasonable. To explore this further, the next section
provides an overview of regularization, and how different forms can be
used to perform variable selection across a broad range of models.

@@

Although a whole host of methods exist to perform variable selection,
the use of regularization has seen a wide array of application in the
context of regression, and more recently, in areas such as graphical
modeling, as well as a host of others. The two most common procedures
for regularization in regression are the \textit{ridge} (Hoerl and
Kennard 1970) and the least absolute shrinkage and selection operator
(\textit{lasso}; Tibshirani 1996); however, there are various
alternative forms that can be seen as subsets or generalizations of
these two procedures. Given an outcome vector \textit{y} and predictor
matrix \(X \in {R}^{n \times p}\) , ridge estimates are defined as

@@

In the context of our example depicted in Figure 1, to use lasso
regression to select among the covariates, the growth model would need
to be reduced to two factor scores, which neglects both the relationship
between both the slope and intercept, reducing both to independent
factor scores. Particularly in models with a greater number of latent
variables, this becomes increasingly problematic. A method that keeps
the model structure, while allowing for penalized estimation of specific
parameters is regularized structural equation modeling (RegSEM;
Jacobucci, Grimm, and McArdle 2016). RegSEM adds a penalty function to
the traditional maximum likelihood estimation (MLE) for structural
equation models (SEMs). The maximum likelihood cost function for SEMs
can be written as

@@

\[
P_{enet}(\theta_{j}) = 0.5(1-\alpha)\| \theta_{j} \|_{2} + \alpha\| \theta_{j} \|_{2}
\]

@@

\subsection{Smoothly Clipped Absolute Deviation
Penalty}\label{smoothly-clipped-absolute-deviation-penalty}

@@

\subsection{Minimax Concave Penalty}\label{minimax-concave-penalty}

A final regularization method developed to produce sparser penalties
than the lasso was the minimax concave penalty (MCP, Zhang 2010):

@@

RegSEM is implemented as the \pkg{regsem} package (Jacobucci 2017) in
the \proglang{R} statistical environment (R Core Team 2017). To estimate
the maximum likelihood fit of the model, \pkg{regsem} uses
\textit{Reticular Action Model} (RAM; J. J. McArdle and McDonald 1984,
McArdle (2005)) notation to derive an implied covariance matrix. The
parameters of each SEM are translated into three matrices: the
\textit{filter} (\textit{F}), the \textit{asymmetric} (\textit{A};
directed paths; e.g.~factor loadings or regressions), and the
\textit{symmetric} (\textit{S}; undirected paths; e.g.~covariances or
variances). See Jacobucci, Grimm, and McArdle (2016) for more detail on
RAM notation.

@@

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & ML & lasso & alasso & SCAD & MCP \\ 
  \hline
c1 -$>$ i & 0.92* & 0.72 & 0.91 & 0.94 & 0.92 \\ 
  c2 -$>$ i & 0.07 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c3 -$>$ i & 0.10 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c4 -$>$ i & 0.07 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c5 -$>$ i & 0.04 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c6 -$>$ i & -0.25 & 0.00 & 0.00 & 0.00 & -0.19 \\ 
  c7 -$>$ i & 0.11 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c8 -$>$ i & -0.13 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c9 -$>$ i & -0.03 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c10 -$>$ i & 0.09 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c1 -$>$ s & 1.18* & 1.09 & 1.22 & 1.24 & 1.24 \\ 
  c2 -$>$ s & 0.29* & 0.19 & 0.28 & 0.35 & 0.35 \\ 
  c3 -$>$ s & 0.18 & 0.09 & 0.00 & 0.00 & 0.00 \\ 
  c4 -$>$ s & -0.08 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c5 -$>$ s & -0.18 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c6 -$>$ s & 0.25* & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c7 -$>$ s & -0.18 & -0.04 & 0.00 & 0.00 & 0.00 \\ 
  c8 -$>$ s & 0.26* & 0.10 & 0.00 & 0.00 & 0.00 \\ 
  c9 -$>$ s & -0.06 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c10 -$>$ s & 0.08 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  BIC & 3465.28 & 3427.46 & 3415.05 & 3414.38 & 3417.20 \\ 
   \hline
\end{tabular}
\caption{Parameter estimates for the final models across five estimation methods. Note that * represent significant parameters at p < .05 for maximum likelihood estimation.}
\end{table}

While every regularization method erroneously set both simulated true
intercept effects as zero (as in maximum likelihood, which accounts for
these errors), both the adaptive lasso and SCAD correctly identified
every true zero effect. The lasso identified two false effects while the
MCP mistakenly identified one. Additionally, the lasso estimation of the
true effects was attentuated in comparison to the other regularization
methods. This is in line with previous research (Fan and Li 2001),
necessitating the use of a two-step relaxed lasso method (Meinshausen
2007, see Jacobucci, Grimm, and McArdle (2016)) As expected given the
small ratio between number of estimated parameters and sample size,
maximum likelihood mistakenly identified 3 false effects as significant.

To compare the performance of each penalization method further,
particularly in the presence of a small parameter to sample size ratio,
a small simulation study was conducted. The same model and effects was
kept, but the sample size was varied to include 80, 200, and 1000 to
demonstrate how maximum likelihood estimation of effects improves as
sample size increases, while each of the regularization methods performs
well regardless of sample size. Each run was replicated 200 times. For
each regularization method, the BIC was used to choose a final model
among the 40 penalty vales. The results are displayed in Table 2.

@@

For false positives, the adaptive lasso demonstrated the best
performance, where the performance of maximum likelihood estimation
leveled off at the 0.05 level at a sample size of 1000 as expected. For
false negatives, lasso penalties demonstrated similar results to maximum
likelihood. This was expected given the tendency of the lasso to
under-penalize small coefficients in comparison to the other
regularization methods. The adaptive lasso and SCAD demonstrated
slightly worse results, however, outside of the MCP, each method made
either zero or near zero errors at a sample size of 1000. The poor
performance of the MCP may be in part due to fixing the \(\gamma\)
penalty to 3.7. Varying this parameter may improve the performance of
the method. In summary, the regularization methods demonstrated an
improvement over maximum likelihood, particularly at small samples, for
a model that had a large number of estimated parameters.

@@

This paper provides an introduction to the \pkg{regsem} package,
outlining the mathematical details of the regularized structural
equation modeling (RegSEM; Jacobucci, Grimm, and McArdle 2016) method
and the usage of the \pkg{regsem} package. RegSEM allows the use of
regularization while keeping the structural equation model intact,
adding penalization directly into the estimation of the model. The
application of RegSEM was detailed using two example models: a latent
growth curve model with 20 predictors of both the latent intercept and
slope, along with a factor analysis with one latent factor. With the
latent growth curve model, the small parameter to sample size ratio
resulted in a larger number of false positives in using maximum
likelihood estimation. In both the simulated example and the small
simulation, the different types of regularization in \pkg{regsem}
demonstrated better false positive and negative rates in comparison to
maxiumum likelihood across sample sizes.

Broadly speaking, there is a growing amount of research into the
integration between data mining methods and latent variable models.
Specifically, beyond RegSEM, this has taken the form of item response
theory and regularization (Sun et al. 2016), pairing both structural
equation models with decision trees (Brandmaier et al. 2013),
exploratory psychological network analaysis (e.g. Epskamp, Rhemtulla,
and Borsboom 2016), along with many others. The amount of pairing
between methods that have generally been housed in separate camps will
only increase into the future. This type of research will be facilitated
by the general upsurge in the creation of open source software that
gives users a general framework to test models. This was the motivation
behind creating the \pkg{regsem} package in that users can estimate
models ranging from simple factor analysis models, to latent
longitudinal models with few to many time points, and finally to models
with a large number of latent and observed variables. The use of
regularization allows for the estimation of much larger structural
equation models than before. Specifically, given that latent variables,
as discussed in this article, are more commonly used in the social and
behavioral sciences, sample sizes are generally not large. To estimate
large models with small sample sizes invites increasing amounts of bias
as demonstrated with the simulated data in this paper. Regularization
can be used to reduce the complexity of the model, thus decreasing both
the bias and variance.

@@

This paper provided a brief overview on the use of the \pkg{regsem}
package as an implementation of regularized structural equation
modeling. Because structural equation modeling encompasses a wide array
of latent variable models, the \pkg{regsem} package was created as a
general package for including different forms of regularization into a
host of latent variable models. RegSEM, and thus the \pkg{regsem}
package, has been evaluated in a wide array of SEM models, including
confirmatory factor analysis (Jacobucci, Grimm, and McArdle 2016),
latent change score models (Jacobucci and Grimm 2017), mediation models
(Serang et al. 2017). Future updates to \pkg{regsem} will focus on
decreasing the computational time of large latent variable models in
order to provide an avenue of testing for researchers collecting larger
and larger datasets. RegSEM is a method that operates at all ends of the
data size spectrum: allowing for a reduction in complexity when the
sample size is small, along with dimension reduction in the presence of
large data (both \(N\) and \(P\)).

@@

\hypertarget{ref-McArdle_1984}{}
McArdle, John J., and Roderick P. McDonald. 1984. ``Some Algebraic
Properties of the Reticular Action Model for Moment Structures.''
\emph{British Journal of Mathematical and Statistical Psychology} 37
(2). Wiley-Blackwell: 234--51.

@@

\hypertarget{ref-serang2017xmed}{}
Serang, Sarfaraz, Ross Jacobucci, Kim Brimhall, and Kevin J Grimm. 2017.
``Exploratory Mediation Analysis via Regularization.'' \emph{Structural
Equation Modeling: A Multidisciplinary Journal}. Taylor \& Francis.

>>>>

\Abstract{
The \pkg{regsem} package in \proglang{R}, an implementation of
regularized structural equation modeling (RegSEM; Jacobucci, Grimm, and
McArdle 2016), was recently developed with the goal of incorporating
various forms of penalized likelihood estimation in a broad array of
structural equations models. The forms of regularization include both
the \textit{ridge} (Hoerl and Kennard 1970) and the least absolute
shrinkage and selection operator (\textit{lasso}; Tibshirani 1996),
along with sparser extensions. RegSEM is particularly useful for
structural equation models that have a small parameter to sample size
ratio, as the addition of penalties can reduce the complexity, thus
reducing the bias of the parameter estimates. The paper covers the
algorithmic details and an overview of the use of \pkg{regsem} with the
application of both factor analysis and latent growth curve models.
}

@@

As a simple running example, Figure 1 depicts a linear latent growth
curve model (e.g. Meredith and Tisak 1990) with four time points and ten
predictors for a simulated dataset. In this, a researcher may want to
test this model, but may only have a relatively small sample size
(e.g.~80). There are 29 estimated parameters in this model, resulting in
a estimated parameter to sample size ratio far below even the most
liberal recommendations (e.g.~10:1 parameters to sample size; Kline
2015). In lieu of finding additional respondents, reducing the number of
parameters estimated is one effective strategy for reducing bias.
Specifically, the 20 estimated regressions from \textit{c1-c10} could be
reduced to a number that makes the ratio of the parameters estimated to
sample size more reasonable. To explore this further, the next section
provides an overview of regularization, and how different forms can be
used to perform variable selection across a broad range of models.

@@

Although a host of methods exist to perform variable selection, the use
of regularization has seen a wide array of application in the context of
regression, and more recently, in areas such as graphical modeling, as
well as a host of others. The two most common procedures for
regularization in regression are the \textit{ridge} (Hoerl and Kennard
1970) and the least absolute shrinkage and selection operator
(\textit{lasso}; Tibshirani 1996); however, there are various
alternative forms that can be seen as subsets or generalizations of
these two procedures. Given an outcome vector \textit{y} and predictor
matrix \(X \in {R}^{n \times p}\) , ridge estimates are defined as

@@

In the context of our example depicted in Figure 1, to use lasso
regression to select among the covariates, the growth model would need
to be reduced to two factor scores, which neglects both the relationship
between both the slope and intercept, reducing both to independent
variables. Particularly in models with a greater number of latent
variables, this becomes increasingly problematic. A method that keeps
the model structure, while allowing for penalized estimation of specific
parameters is regularized structural equation modeling (RegSEM;
Jacobucci, Grimm, and McArdle 2016). RegSEM adds a penalty function to
the traditional maximum likelihood estimation (MLE) for structural
equation models (SEMs). The maximum likelihood cost function for SEMs
can be written as

@@

\[
P_{enet}(\theta_{j}) = (1-\alpha)\| \theta_{j} \|_{2} + \alpha\| \theta_{j} \|_{1}
\]

@@

\subsection{Sparse Extensionse}\label{sparse-extensionse}

@@

Additionally, the MCP takes the form of:

@@

RegSEM is implemented as the \pkg{regsem} package (Jacobucci 2017) in
the \proglang{R} statistical environment (R Core Team 2017). To estimate
the maximum likelihood fit of the model, \pkg{regsem} uses
\textit{Reticular Action Model} (RAM; McArdle and McDonald 1984; McArdle
2005) notation to derive an implied covariance matrix. The parameters of
each SEM are translated into three matrices: the \textit{filter}
(\textit{F}), the \textit{asymmetric} (\textit{A}; directed paths;
e.g.~factor loadings or regressions), and the \textit{symmetric}
(\textit{S}; undirected paths; e.g.~covariances or variances). See
Jacobucci, Grimm, and McArdle (2016) for more detail on RAM notation and
its application to RegsEm.

@@

Structural equation modeling is hard, and pairing with regularization
doesn't make it any easier. Given this, and the number of options
available in the \pkg{regsem} package, a Google group forum was created
in order to answer questions and trouble shoot at
\url{https://groups.google.com/forum/\#!forum/regsem}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & MLE & lasso & alasso & SCAD & MCP \\ 
  \hline
c1 -$>$ i & 0.92* & 0.72 & 0.91 & 0.94 & 0.92 \\ 
  c2 -$>$ i & 0.07 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c3 -$>$ i & 0.10 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c4 -$>$ i & 0.07 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c5 -$>$ i & 0.04 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c6 -$>$ i & -0.25 & 0.00 & 0.00 & 0.00 & -0.19 \\ 
  c7 -$>$ i & 0.11 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c8 -$>$ i & -0.13 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c9 -$>$ i & -0.03 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c10 -$>$ i & 0.09 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c1 -$>$ s & 1.18* & 1.09 & 1.22 & 1.24 & 1.24 \\ 
  c2 -$>$ s & 0.29* & 0.19 & 0.28 & 0.35 & 0.35 \\ 
  c3 -$>$ s & 0.18 & 0.09 & 0.00 & 0.00 & 0.00 \\ 
  c4 -$>$ s & -0.08 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c5 -$>$ s & -0.18 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c6 -$>$ s & 0.25* & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c7 -$>$ s & -0.18 & -0.04 & 0.00 & 0.00 & 0.00 \\ 
  c8 -$>$ s & 0.26* & 0.10 & 0.00 & 0.00 & 0.00 \\ 
  c9 -$>$ s & -0.06 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  c10 -$>$ s & 0.08 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  BIC & 3465.28 & 3427.46 & 3415.05 & 3414.38 & 3417.20 \\ 
   \hline
\end{tabular}
\caption{Parameter estimates for the final models across five estimation methods. Note that * represent significant parameters at p < .05 for maximum likelihood estimation.}
\end{table}

While every regularization method erroneously set both simulated true
intercept effects as zero (non-significant in MLE), both the adaptive
lasso and SCAD correctly identified every true zero effect. The lasso
identified two false effects while the MCP mistakenly identified one.
Additionally, the lasso estimation of the true effects was attentuated
in comparison to the other regularization methods. This is in line with
previous research (Fan and Li 2001), necessitating the use of a two-step
relaxed lasso method (Meinshausen 2007; see Jacobucci, Grimm, and
McArdle 2016) As expected given the small ratio between number of
estimated parameters and sample size, MLE mistakenly identified 3 false
effects as significant.

To compare the performance of each penalization method further,
particularly in the presence of a small parameter to sample size ratio,
a small simulation study was conducted. The same model and effects was
kept, but the sample size was varied to include 80, 200, and 1000 to
demonstrate how MLE improves as sample size increases, while each of the
regularization methods performs well regardless of sample size. Each run
was replicated 200 times. For each regularization method, the BIC was
used to choose a final model among the 40 penalty vales. The results are
displayed in Table 2.

@@

For false positives, the adaptive lasso demonstrated the best
performance, where the performance of MLEleveled off at the 0.05 level
at a sample size of 1000 as expected. For false negatives, lasso
penalties demonstrated similar results to MLE. This was expected given
the tendency of the lasso to under-penalize small coefficients in
comparison to the other regularization methods. The adaptive lasso and
SCAD demonstrated slightly worse results, however, outside of the MCP,
each method made either zero or near zero errors at a sample size of
1000. The poor performance of the MCP may be in part due to fixing the
\(\gamma\) penalty to 3.7. Varying this parameter may improve the
performance of the method. In summary, the regularization methods
demonstrated an improvement over maximum likelihood, particularly at
small samples, for a model that had a large number of estimated
parameters.

@@

This paper provides an introduction to the \pkg{regsem} package,
outlining the mathematical details of regularized structural equation
modeling (RegSEM; Jacobucci, Grimm, and McArdle 2016) and the usage of
the \pkg{regsem} package. RegSEM allows the use of regularization while
keeping the structural equation model intact, adding penalization
directly into the estimation of the model. The application of RegSEM was
detailed using two example models: a latent growth curve model with 20
predictors of both the latent intercept and slope, along with a factor
analysis with one latent factor. With the latent growth curve model, the
small parameter to sample size ratio resulted in a larger number of
false positives in using maximum likelihood estimation. In both the
simulated example and the small simulation, the different types of
regularization in \pkg{regsem} demonstrated better false positive and
negative rates in comparison to maxiumum likelihood across sample sizes.

Broadly speaking, there is a growing amount of research into the
integration between data mining methods and latent variable models.
Specifically, beyond RegSEM, this has taken the form of item response
theory and regularization (Sun et al. 2016), pairing both structural
equation models with decision trees (Brandmaier et al. 2013),
exploratory psychological network analaysis (e.g. Epskamp, Rhemtulla,
and Borsboom 2016), along with many others. The amount of pairing
between methods that have generally been housed in separate camps will
only increase into the future. This type of research will be facilitated
by the general upsurge in the creation of open source software that
gives users a general framework to test models. This was the motivation
behind creating the \pkg{regsem} package, in that users can estimate
models ranging from simple factor analysis models, to latent
longitudinal models with few to many time points, and finally to models
with a large number of latent and observed variables. The use of
regularization allows for the estimation of much larger structural
equation models than before. Specifically, given that latent variables,
as discussed in this article, are more commonly used in the social and
behavioral sciences, sample sizes are generally not large. To estimate
large models with small sample sizes invites increasing amounts of bias
as demonstrated with the simulated data in this paper. Regularization
can be used to reduce the complexity of the model, thus decreasing both
the bias and variance.

@@

This paper provided a brief overview on the use of the \pkg{regsem}
package as an implementation of regularized structural equation
modeling. Because structural equation modeling encompasses a wide array
of latent variable models, the \pkg{regsem} package was created as a
general package for including different forms of regularization into a
host of latent variable models. RegSEM, and thus the \pkg{regsem}
package, has been evaluated in a wide array of SEM models, including
confirmatory factor analysis (Jacobucci, Grimm, and McArdle 2016),
latent change score models (Jacobucci and Grimm 2017), mediation models
(Serang et al. in press). Future updates to \pkg{regsem} will focus on
decreasing the computational time of large latent variable models in
order to provide an avenue of testing for researchers collecting larger
and larger datasets. RegSEM is a method that operates at all ends of the
data size spectrum: allowing for a reduction in complexity when the
sample size is small, along with dimension reduction in the presence of
large data (both \(N\) and \(P\)).

@@

\hypertarget{ref-McArdle_1984}{}
McArdle, John J, and Roderick P. McDonald. 1984. ``Some Algebraic
Properties of the Reticular Action Model for Moment Structures.''
\emph{British Journal of Mathematical and Statistical Psychology} 37
(2). Wiley-Blackwell: 234--51.

@@

\hypertarget{ref-serang2017xmed}{}
Serang, Sarfaraz, Ross Jacobucci, Kim Brimhall, and Kevin J Grimm. in
press. ``Exploratory Mediation Analysis via Regularization.''
\emph{Structural Equation Modeling: A Multidisciplinary Journal}. Taylor
\& Francis.
